---
title: "Predicting maximal growth rates with `gRodon`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{gRodon-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The `gRodon` package allows you to predict the maximal growth rate of
a prokaryotic organism from genomic data. For details of how this
is done please see the original paper. Briefly, gRodon exploits codon usage
statistics to detect optimization of highly-expressed genes, an indicator of
selection for rapid growth. We will provide some guidance about when these predictions are 
appropriate below. 

The `gRodon` package is quite simple, with only a single function currently
available to users: `predictGrowth()`. We provide a simple, easy to
use interface that allows any user with a genome in-hand to predict maximal
growth rate. 

The `gRodon` approach is largely based on an earlier program, `growthpred`,
developed by [Vieira-Silva et al.](https://doi.org/10.1371/journal.pgen.1000808). 
`gRodon` is intended to be:  
1. User-friendly (especially for R users)  
2. More accurate (due to the incorporation of additional measures of codon usage)  
3. More accurate on mixed-organism samples (i.e., metagenomes; due to a correction factor for the relative abundance of different organisms)

## Installation

The easiest way to install gRodon is with [devtools](https://github.com/r-lib/devtools):

```{r devtools, eval=FALSE}
devtools::install_github("jlw-ecoevo/gRodon")
```

`gRodon` has a couple dependencies - namely the `Biostrings` and `coRdon`
packages which are bioconductor packages and cannot be installed via CRAN. 
To install them run the following:

```{r dependencies, eval=FALSE }
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("Biostrings")
BiocManager::install("coRdon")
```

## Before You Begin

There are two things that `gRodon` does *not* do that you *must* do ahead of 
time yourself in order to predict growth rate:  
1. You must identify coding sequences in your genome  
(e.g., using [prodigal](https://github.com/hyattpd/Prodigal))  
2. You must annotate a set of highly-expressed genes, typically those coding for ribosomal 
proteins. We usually use genes coding for ribosomal proteins as our highly-expressed set since
these are universal across microbes and almost always highly expressed (if you wanna
grow you gotta make proteins), but if you have expression data for your organism
you might be able to tailor an even better set of highly expressed genes.

It is helpful if (in frame) coding sequences are stored in a fasta file for 
easy loading. Additionally, you will need to have a logical vector describing
which genes are part of your highly expressed set (this should become a bit
clearer with the examples below).

##### A Tiny Detour: Annotating Your Assembly

I'll lead you through one way to get annotated coding sequence to input into gRodon
here, but there are many (MANY) ways to do this. For example, check out this set of
tutorials on [functional annotation](https://github.com/biovcnet/biovcnet.github.io/wiki/TOPIC%3A-Functional-Annotation).

**Disclaimer:** *I make no claim that this is the best, most efficient, or most graceful way
to get annotated coding sequences from your assembly, but it will get the job done.
If you have strong opinions about how this should be done shoot me an email and I
will be happy to link to your rant/tutorial. If you already have annotated coding sequences please skip this section. If you have unannotated CDS (e.g., from [prodigal](https://github.com/hyattpd/Prodigal)),
you might consider simply BLASTing your output for highly expressed genes (e.g., using
the blastdb  of ribosomal proteins included in [growthpred](http://ftp.pasteur.fr/pub/gensoft/projects/growthpred/)).*

Given an un-annotated genome or metagenomic assembly, an easy way
to find and annotate coding sequence is using the program [prokka](https://github.com/tseemann/prokka).
For example, if I have a set of assembled contigs, I could run (on the command line):
```
prokka contig.fasta --prefix my_genome
```
This will give several output files. You are interested the untranslated coding sequences (CDS), which 
can be found in the `*.ffn` file along with other non-coding transcripts (e.g., rRNA, tRNA). To get the names
of just your CDS you can pull them out of the `*.gff` output:
```
sed -n '/##FASTA/q;p' my_genome/my_genome.gff | awk '$3=="CDS"' | awk '{print $9'} | awk 'gsub(";.*","")' | awk 'gsub("ID=","")' > CDS_names.txt
```
The above command first trims the sequence data off the end of the `*.gff` file, then looks for
rows describing CDS and pulls out their IDs, finally it outputs these IDs to 
a file called `CDS_names.txt`.

```{r prokka, eval=FALSE}
# Load your *.ffn file into R
genes <- readDNAStringSet("my_genome/my_genome.ffn")

# Subset your sequences to those that code for proteins
CDS_IDs <- readLines("CDS_names.txt")
gene_IDs <- gsub(" .*","",names(genes)) #Just look at first part of name before the space
genes <- genes[gene_IDs %in% CDS_IDs]

#Search for genes annotated as ribosomal proteins
highly_expressed <- grepl("ribosomal protein",names(genes),ignore.case = T)
```

You are now ready to run gRodon.

## A Minimal Example

We have included an example genome assembly with this package, downloaded 
from NCBI's RefSeq database (*Streptococcus pyogenes* M1, GCF_000349925.2). This
is simply a fasta file with the predicted coding sequences and 
annotations for this genome (provided by NCBI).

Let's load this file using the `readDNAStringSet()` function from the
`Biostrings` package (required for `gRodon` to work).

```{r setup, message=FALSE}
library(gRodon)
library(Biostrings)
```

```{r load}
path_to_genome <- system.file('extdata',
  'GCF_000349925.2_ASM34992v2_cds_from_genomic.fna.gz',
  package = 'gRodon')
genes <- readDNAStringSet(path_to_genome)
```


We also need a set of highly expressed genes. In general, the ribosomal
proteins are a good set of genes to use for this purpose. Since these proteins
should already be annotated in our example file (try running `names(genes)` to
see the annotations from NCBI) we  can use `grep()` to search for them 
(specifically, `grepl()` to return a logical vector).

```{r HE}
highly_expressed <- grepl("ribosomal protein",names(genes),ignore.case = T)
```

And now we are ready to predict the maximal growth rate of *S. pyogenes* M1.

```{r Pred}
predictGrowth(genes, highly_expressed)
```

The warning at the top is letting you know that there were a number of genes
in your fasta that either weren't a length multiple of three, or were under the 
length threshold of 80 codons (240 bp) where estimates of bias will be unreliable. Don't
worry about this unless the number of filtered genes is very high compared
to the total number of genes in your file. We expect some genes to typically get
filtered.

The output contains several (hopefully) useful quantities:

*  `$CUBHE` is the mean of the codon usage bias of each highly expressed gene relative to
all other genes
*  `$ConsistencyHE` is the mean of the codon usage bias of each highly expressed gene relative to
all other highly expressed genes
*  `$CPB` is the genome-wide codon pair bias
*  `$FilteredSequences` is the number of sequences filtered due to length
*  `$d` is the estimated minimal doubling time from `gRodon` in hours. This is, presumably, what
you came here for.
*  `$LowerCI` and `$UpperCI` are the 95% confidence intervals for `d`

For details on how each of these is calculated please see the `gRodon` paper. For 
most users the only outputs that will matter much to you are `$d`, `$LowerCI`, and `$UpperCI`, which are your
estimated minimal doubling times and 95% CIs output by the gRodon model.

For the most part, that's all you need to know. There are a few specific use
cases we discuss below that require a slightly more thoughful application of
`gRodon`. These are:

*  The use of partial genomic data (e.g., incomplete SAGs or MAGs)  
*  The use of metagenomic data  (**if using metagenomes you MUST use metagenome mode or risk getting some very strange results**)
*  Predicting maximal growth rates when maximal growth is very slow  
*  Predicting the maximal growth rate of a psychrophile or thermophile  

### Partial Mode

What if you don't have a nice, complete genome ... can you still use `gRodon`? 
Yes, you can, though you may want to use either "partial" or "metagenome" mode.

When your genome is incomplete, you may have insufficient data to accurately
estimate the codon pair bias (since there are many possible codon pairs). In 
this case you can set `gRodon` to "partial" mode, which excludes pair-bias 
from the prediction. This is probably a good choice when working with incomplete
SAGs and MAGs. The expected decrease in accuracy is quite small (see 
`gRodon` paper).

```{r PredP}
predictGrowth(genes, highly_expressed, mode="partial")
```

### Metagenome Mode

For bulk estimation of the average community-wide maximal growth rate from metagenomes, 
our consistency statistic is not appropriate, since different
organisms may prefer different codons (even when they have similar bias values).
It also doesn't make much sense to calculate the pair-bias in this scenario.
Thus the model for predicting the average maximal growth rate of a metagenomic sample
excludes consistency and pair-bias. Importantly, metagenome mode is expected to
be less accurate than the default mode, so only use this mode if you must 
(i.e., you have a metagenomic sample).

Let's try this on one of the samples from [Okie et al.](https://doi.org/10.7554/eLife.49816), as discussed in the
`gRodon` manuscript:

```{r PredM}
path_to_metagenome <- system.file('extdata',
  'ERR2143764_fastp_prokka_scaffolds.ffn.gz',
  package = 'gRodon')
genes <- readDNAStringSet(path_to_metagenome)
highly_expressed <- grepl("ribosomal protein",names(genes),ignore.case = T)
predictGrowth(genes, highly_expressed, mode = "metagenome")
```

You may also want to incorporate a correction for the relative abundance of different
organisms in this calculation (e.g., weighted metagenome mode) by specifying the `depth_of_coverage`
option. This should lead to more realistic estimates of the average community-wide maximal growth rate.
To use this correction provide `depth_of_coverage` as a vector of mean depths of coverage for each gene.
It does not matter if these coverages are normalized since we only care about relative coverage.

First let's take a look at our read depths:

```{r readdepths}
path_to_coverage <- system.file('extdata',
  'ERR2143764_fastp_map2ffn_counts.tsv',
  package = 'gRodon')
read_depths <- read.delim(path_to_coverage,
                            stringsAsFactors = FALSE)
depths <- read_depths$meandepth
#Make sure in the correct order
names(depths) <- read_depths$X.rname
depth_of_coverage <- depths[gsub(" .*", "", names(genes))]
head(depth_of_coverage)
```

And now let's run weighted metagenome mode:

```{r weightedM}
predictGrowth(genes, 
              highly_expressed, 
              mode = "metagenome", 
              depth_of_coverage = depth_of_coverage)
```

### Slow Growers

All codon-usage based predictors of maximal growth will underestimate minimal doubling times
when these doubling times are very long. In other words, these predictors will often 
indicate that slow-growing microbes are able to grow more quickly than they actually can. 
See the original paper for a more in-depth discusison, but this is because 
codon-usage bias tends to plateau at a lower-bound in slow-growers. 

What does this mean for you? Well, `gRodon` can reliably tell you if your
microbe grows slowly, but it can't tell you how slowly. In practice, minimal doubling 
time predictions appear to be accurate up to 5 hours. If `gRodon` predicts a
doubling time above 5 hours, you can confidently say your microbe grows slowly,
but not quite how slowly (minimal doubling times >5hrs tend to be underestimates). 
In these cases `gRodon` will even warn you to be  careful:

```{r slow}
path_to_genome <- system.file('extdata',
  'GCF_003253775.1_ASM325377v1_cds_from_genomic.fna.gz',
  package = 'gRodon')
genes <- readDNAStringSet(path_to_genome)
highly_expressed <- grepl("ribosomal protein",names(genes),ignore.case = T)
predictGrowth(genes, highly_expressed)
```

The example above is for a *Mycobacterium leprae* genome. This organism has a 
doubling time around 240 hours. So we are off by an order of magnitude, but 
still well within the "slow grower" range.

### Psychrophiles and Thermophiles

[Vieira-Silva et al.](https://doi.org/10.1371/journal.pgen.1000808) showed that
codon-usage based maximal growth estimators are sensitive to growth temperature. 
Thermophiles and psychrophiles showed consistently higher/lower growth rates 
respectively. We included a temperature correstion factor in `gRodon` to account
for this. We caution that this correction is based on only 31 extremophile
species, and that `gRodon` has primarily been validated on 
mesophilic organisms.

```{r temp}
predictGrowth(genes, highly_expressed, temperature = 33)
```

If working with thermophiles or psychrophiles you might consider using `gRodon`'s
alternative training set (below) which includes more extremophiles than
the default training set.

### Alternative Training Set from [Madin et al.](https://doi.org/10.1038/s41597-020-0497-4)

We also include a version of `gRodon` trained on an alternative set of minimal
doubling time estimates from a trait database compiled by [Madin et al.](https://doi.org/10.1038/s41597-020-0497-4).  While this is a larger training
set than the set of minimal doubling times fron [Vieira-Silva et al.](https://doi.org/10.1371/journal.pgen.1000808), it is not hand-curated in the
same way and is likely to be more error-prone. In any case, both training sets give
very similar results. See the `gRodon` paper for an in-depth discussion.


```{r madin}
path_to_genome <- system.file('extdata',
  'GCF_000349925.2_ASM34992v2_cds_from_genomic.fna.gz',
  package = 'gRodon')
genes <- readDNAStringSet(path_to_genome)
highly_expressed <- grepl("ribosomal protein",names(genes),ignore.case = T)
predictGrowth(genes, highly_expressed, training_set = "madin")
```

